# Matrices in R
**Created by John Palowitch, UNC Chapel Hill**

**Edited by Robin Cunningham, UNC Chapel Hill**


### Conceptual Questions 
Please read the document ***Matrices in R.docx*** before completing this assignment. This assignment has two goals:

-Learning new commands in R from a written source.

-Using R for Matrix Calculations
<br><br>
1. Suppose:
  \[
  \mathbf{A} = \begin{bmatrix}2&3&4&5\\-1&0&3&2\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}4&-1&10&0\\-2&-2&5&6\end{bmatrix}
  \]
  \[
  \mathbf{a} = \begin{bmatrix}6&2&-11&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}4\\2\\3\\0\end{bmatrix},\;\;\;\;\mathbf{c} = \begin{bmatrix}-3\\-10\end{bmatrix}
  \]
    a. On this page, write down the dimensions of the following (on this sheet, next to the expression). If the operation is not possible, write "N.P."
        <br>i. $\mathbf{A}^T\mathbf{B}$   4X4 (first one done for you!)
        <br>ii. $\mathbf{B}\mathbf{A}^T$ 2X2
        <br>iii. $\mathbf{A}\mathbf{B}$ N.P.
        <br>iv. $\mathbf{A}^T\mathbf{A}$ 4X4
        <br>v. $\mathbf{b}^T\mathbf{A}$ N.P.
        <br>vi. $\mathbf{a}\mathbf{A}^T$ 1X2
        <br>vii. $\mathbf{B}\mathbf{b}$ 2X1
        <br>viii. $\mathbf{B}\mathbf{c}^T$ N.P.
        <br>ix. $\mathbf{a}\mathbf{c}$ N.P.
        <br>x. $\mathbf{b}\mathbf{a}$ 4X4
    <br><br>b. Again, write down the dimensions of the following (on this sheet, next to the expression). If the operation is not possible, write "N.P."
        <br>i. $(\mathbf{A}^T\mathbf{B})^{-1}\mathbf{a}$ N.P.
        <br>ii. $\mathbf{c}(\mathbf{B}\mathbf{A}^T)^{-1}$ N.P.
        <br>iii. $(\mathbf{B}^T\mathbf{B})^{-1}\mathbf{B}^T\mathbf{c}$ N.P.
        <br>iv. $\mathbf{b}^T\mathbf{A}^T\mathbf{c}$ 1X1
        <br>v. $(\mathbf{B}\mathbf{B}^T)^{-1}\mathbf{B}\mathbf{b}$ 2x1
      
        
### Applied Questions 
2. In a code block below, compute and print the following matrices and vectors:
\[
\mathbf{A} = \begin{bmatrix}-6&0&1\\1&10&7.5\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}-15&1&1\\5.25&1&2.5\end{bmatrix}
\]
\[
\mathbf{a} = \begin{bmatrix}6&2&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}2\\-7\end{bmatrix}
\]
```{r, eval=TRUE}
A = matrix(c(-6, 1, 0, 10, 1, 7.5), nrow=2, ncol=3)
B = matrix(c(-15, 5.25, 1, 1, 1, 2.5), nrow=2, ncol=3)
a = matrix(c(6, 2, 1), nrow=1, ncol=3)
b = matrix(c(2, -7), nrow=2, ncol=1)
A
B
a
b
```

3. Compute (with **R**, in a new code block below) and print the following products:
    a. $\mathbf{A}^T\mathbf{B}$
    b. $\mathbf{B}\mathbf{a}^T$
    c. $\mathbf{b}^T\mathbf{A}$
    d. $\mathbf{A}^T\mathbf{b}$
    e. $\mathbf{a}^T\mathbf{a}$
    f. $\mathbf{a}\mathbf{a}^T$
    g. $(\mathbf{A}^T\mathbf{A})^{-1}$
    h. $(\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{b}$
    i. $\mathbf{b}^T\mathbf{A}\mathbf{a}^T$
    j. $\mathbf{a}\mathbf{B}\mathbf{A}^T\mathbf{b}$
    ```{r, eval=TRUE}
    t(A) %*% B
    B %*% t(a)
    t(b) %*% A
    t(A) %*% b
    t(a) %*% a
    a %*% t(a)
    #solve(t(A) %*% A)
    #solve(t(A) %*% A) %*% t(A) %*% b
    t(b) %*% A %*% t(a)
    #a %*% B %*% t(A) %*% b
    ```


